---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```
LWFBrook90R: Run the LWF-BROOK90 hydrological model from within R.

# Motivation
In hydrology, many R-packages exist that deal with pre- and post-processing of input data and results of hydrological process models. In addition, many ready-to-use algorithms exist in R providing automatic calibration, sensitivity analysis, and parallelisation techniques. In order to make the vast resources of R directly available to the 1D-SVAT model [LWF-BROOK90](https://www.lwf.bayern.de/boden-klima/wasserhaushalt/index.php), *LWFBrook90R* was developed. The Fortran model code is integrated into the package as dynamic library and compiles, when installing *LWFBrook90R*. 
Basically, *LWFBrook90R* features the following functionality in one function call to the package core function `runLWFB90()`:

* create model input objects from climate driving data, model control options and parameters,
* execute the LWF-BROOK90 model code,
* read and return the created output files.

The model control options thereby let you select different functions for defining aboveground stand dynamics, phenology, and root length density depth distributions. Additionally, a set of convenience functions are available to set up the required lists of model control options and parameters, and to derive soil hydraulic parameters from soil physical properties using pedotransfer-functions. Furthermore, wrapper-functions for `runLWFB90()` are, available facilitating parallel multi-run simulations and multi-site simulations.

# Installation

Before installing the *LWFBrook90R* R-package, be sure to download and install the following packages:
```{r, eval = F}
install.packages("data.table")
install.packages("vegperiod")
install.packages("sirad")
install.packages("foreach")
install.packages("doSNOW")
```

Now you can install *LWFBrook90R*. The package is available on GitHub and can be installed using the *devtools*-package:
```{r, eval = F}
if (!requireNamespace("devtools")) {
    install.packages("devtools")
  }
devtools::install_github(repo = "pschmidtwalter/LWFBrook90R")
```

# Basic usage
Load LWFBrook90R
```{r, eval = F}
library(LWFBrook90R)
```

Load sample meteo and soil data
```{r, eval = F}
data("slb1_meteo")
data("slb1_soil")
```

Set up lists containing default model control options and model parameters
```{r, eval = F}
options.b90 <- setoptions_LWFB90()
param.b90 <- setparam_LWFB90()
```
Set new start and end dates in model control options
```{r, eval = F}
options.b90$startdate <- as.Date("2002-01-01")
options.b90$enddate <- as.Date("2003-12-31")
```

Derive soil hydraulic properties from soil physical properties using a pedotransfer function
```{r, eval = F}
soil <- cbind(slb1_soil, hydpar_puh2(clay = slb1_soil$clay,
                                     silt = slb1_soil$silt,
                                     sand = slb1_soil$sand,
                                     bd = slb1_soil$bd,
                                     oc.pct = slb1_soil$c_org))
```

Run LWF-Brook90 with the created input and store the results in b90.results.slb1
```{r, results = 'hide', eval =F}
b90.results.slb1 <- runLWFB90(project.dir = "example_run_b90/",
                            param.b90 = param.b90,
                            options.b90 = options.b90,
                            soil = soil,
                            climate = slb1_meteo)
```

# Status

## R-Code

The package works as intended and is fully documented. However, there are some points to be accomplish in the near future:

- [x] enable use of Clapp-Horneberg hydraulics in addition to the default Mualem-van Genuchten
- [ ] Run the `check` with Travis.
- [ ] implement Goodness-of-fit measures with respect to observations that can be returned on top / instead of actual simulation results.

## Fortran-Code 
 
- [x] Use of sub-day resolution precipitation interval data.
- [x] Model output results tested against the output from the original b90.exe commandline tool. 
- [x] Cleaning up `declared but not used` variables
- [x] Making the writing of model log-file that contains the former commandline-feed optional. 

# Authors 
Paul Schmidt-Walter & Volodymyr Trotsiuk

# License
GPL-3 for the package. License for LWF-BROOK90 is unknown.
