---
title: "Multi-Site simulations using `run_multisite_LWFB90()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Multi-Site simulations using `run_multisite_LWFB90()`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

In the previous vignette ['Multi-run simulations in LWFBrook90R'](LWFBrook90R-3-Multiruns.html), we learned how to make multiple simulations using a set of variable model parameters using the function `run_multi_LWFB90()`. 
To simulate a set of different sites with different soil, climate and vegetation input, we can use the function `run_multisite_LWFB90()` that is the subject of this vignette.

```{r, warning = FALSE, message = FALSE, eval = TRUE}
library(LWFBrook90R)
library(data.table)
data("slb1_meteo")
data("slb1_soil")
soil <- cbind(slb1_soil, hydpar_wessolek_tab(texture = slb1_soil$texture))
```

## List input for `soil`, `climate` and `param_b90`

The function `run_multisite_LWFB90()` runs through lists of `param_b90`, `climate`, and `soil`-objects, and evaluates the specified parameter sets for each of the soil/climate combinations. To
demonstrate its usage, we define two parameter sets, that we want to run on
three different sites (i.e. unique combinations of climate and soil). We include the two parameter sets in a list `parms_l`:

```{r}
parms_beech <- set_paramLWFB90(maxlai = 6)
parms_spruce <- set_paramLWFB90(maxlai = 4.5, winlaifrac = 0.8)
parms_l <- list(beech = parms_beech, spruce = parms_spruce)
```

We pretend that the three sites all have individual climates and soils, and set
up lists for soil and climate input:

```{r}
soils_l <- list(soil1 = soil, soil2 = soil, soil3 = soil)
climates_l <- list(clim1 = slb1_meteo, clim2 = slb1_meteo, clim3 = slb1_meteo)
```

Now we can run a small example:

```{r}
startdate <- as.Date("2002-06-01")
enddate <- as.Date("2002-06-30")

msite_run1 <- run_multisite_LWFB90(
  options_b90 = set_optionsLWFB90(startdate = startdate, enddate = enddate),
  param_b90 = parms_l,
  climate = climates_l,
  soil = soils_l,
  cores = 2)
```

The results are returned as a named list of single run objects, with their names being
concatenated from the names of the input list entries holding the individual
`param_b90`, `climate`, and `soil` input objects:

```{r}
str(msite_run1, max.level = 1)
```


## Data management (ii): A function as `climate`-argument

The function `run_multisite_LWFB90()` can easily be set up to run a few dozens
of sites with individual climate data. However, simulating thousands of sites
can easily cause errors, because such a large list of `climate` data.frames
might overload the memory of a usual desktop computer. Fortunately, it is
possible to pass a function instead of a data.frame as `climate`-argument to
`run_LWFB90()`. Such a function can be used to create the `climate`-data.frame
from a file or database-connection within `run_LWFB90()` or
`run_multisite_LWFB90()` on the fly. 

For `run_LWFB90()`, we can simply provide arguments to the function via the
`...`-placeholder. For `run_multisite_LWFB90()`, we need to pass arguments to a
`climate`-function (possibly with individual values for individual site, e.g. a
file name) via the `climate_args`-argument.

To demonstrate this mechanism, we write three files with climatic data to a
temporary location, from where we will read them back in later:

```{r, result='hide'}
tdir <- tempdir()
fnames <- paste0(tdir, "/clim", 1:3, ".csv")
lapply(fnames, function(x) {
  write.csv(slb1_meteo[year(slb1_meteo$dates) == 2002,], 
            file = x, row.names = FALSE)
})

```

For testing, we perform a single run with `run_LWFB90()` and use the `fread`
function from the 'data.table'-package as `climate`-argument. The function reads
text-files, and takes a `file` name as argument that we include in the call. It
points to the first of our three climate files:

```{r}
srun <- run_LWFB90(
  options_b90 = set_optionsLWFB90(startdate = startdate, enddate = enddate),
  param_b90 = set_paramLWFB90(),
  soil = soil,
  climate = fread,
  file = fnames[1],
  rtrn.input = FALSE)
```

The same construct basically works with the function `run_multisite_LWFB90()`.
The only difference to single-run simulations is that the arguments for the
function have to be specified in a named list of lists, one sub-list for each
site. We set it up as follows:

```{r}
clim_args <- list(climfromfile1 = list(file = fnames[1]),
                  climfromfile2 = list(file = fnames[2]),
                  climfromfile3 = list(file = fnames[3]))
```

Now we call `run_multisite_LWFB90()`, and set up the function `fread` as
`climate`-parameter. Our list of lists with individual arguments for `fread` is
passed to the function via `climate_args`:

```{r}
msite_run2 <- run_multisite_LWFB90(
  options_b90 = set_optionsLWFB90(startdate = startdate, enddate = enddate),
  param_b90 = parms_l,
  soil = soils_l,
  climate = fread,
  climate_args = clim_args,
  cores = 2)
```

We simulated two parameter sets using three different climate/soil combinations:

```{r}
str(msite_run2, max.level = 1)
```

The names of the climate used in the result names are now coming from the
top-level names of our list `clim_args`, because we used a function as
`climate`-argument. The function `fread` is evaluated directly within
`run_multisite_LWFB90()`, and is not passed to `run_LWFB90()`, because otherwise
it would have been evaluated for each single-run simulation. In this way,
`fread` is evaluated only three times for in total six simulations which saves
us some execution time, in case we want to simulate multiple parameter sets using the
same climatic data.

## Multi-site simulation: Input from file, output to file

Now that we learned how to use a function as climate input, we can combine this
input facility with an `output_fun` that writes the simulation results to a
file. To do so, we extend our output function from the previous vignette
['Multi-run simulations in LWFBrook90R'](LWFBrook90R-3-Multiruns.html) so that
it writes the aggregated results to a file in a specified directory. The file name is
constructed from the names of the current soil, climate, and parameter object,
which are passed automatically from `run_multisite_LWFB90()` to `run_LWFB90()`
as character variables `soil_nm`, `clim_nm`, and `param_nm`. In this way, the
names of currently processed input objects are accessible to
`output_fun`-functions within `run_LWFB90()`.

```{r}
output_function <- function(x, tolayer, basedir = getwd(),
                            soil_nm, clim_nm, param_nm ) {
  # file-name
  filenm = file.path(basedir, paste(soil_nm, clim_nm, param_nm, sep = "_"))
  
  # aggregate SWAT
  swat_tran <- x$layer_output[which(nl <= tolayer), 
                             list(swat = sum(swati)),
                             by  = list(yr, doy)]
  #add transpiration from EVAPDAY.ASC
  swat_tran$tran <- x$output$tran
  
  # get beginning and end of growing season from input parameters
  vpstart <- x$model_input$param_b90$budburstdoy
  vpend <- x$model_input$param_b90$leaffalldoy
  swat_tran <- merge(swat_tran,
                     data.frame(yr = unique(swat_tran$yr),
                                vpstart, vpend), by = "yr")
  # mean swat and tran sum
  swattran_vp <- swat_tran[doy >= vpstart & doy <= vpend, 
            list(swat_vp_mean = mean(swat), tran_vp_sum = sum(tran)), by = yr]
  
  write.csv(swattran_vp, file = paste0(filenm, ".csv"))
}
```

Now we can run the simulations, with climate data coming from files, and the
results being written to file our temporary directory `tdir`:

```{r}
msite_run3 <- run_multisite_LWFB90(
  options_b90 = set_optionsLWFB90(startdate = startdate, enddate = enddate),
  param_b90 = parms_l,
  soil = soils_l,
  climate = fread,
  climate_args = clim_args,
  rtrn_input = FALSE, rtrn_output = FALSE,
  output_fun = output_function,
  tolayer = 15,
  basedir = tdir,
  cores = 2)
```

After the simulation has finished, we can list the files and see that our
attempt was successful:

```{r}
list.files(tdir, pattern = "csv")
```

We can also use database connection objects instead of files to read
climate data and save simulation results. For the input of climate data,
connection objects can be defined in advance, and passed directly to the
`climate`-function. However, this does not work for `output_fun` in a parallel
setting like in `run_multisite_LWFB90()` or `run_multi_LWFB90()`, because file
or database connections in R are not exported to parallel workers. Connections
therefore have to be set up (and closed again) within an `output_fun`-function.

```{r, echo = FALSE, results='hide'}
file.remove(list.files(tdir, pattern = "csv", full.names = T))
```
