---
title: "Multi-run simulations in LWFBrook90R"
output: 
  rmarkdown::html_vignette:
    toc: true
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Multi-run simulations in LWFBrook90R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```


## Introduction

With 'LWFBrook90R', parallelized multi-run simulations can be performed
conveniently, extending the basic single-run applications using the function
`run_LWFB90()` described in the introductory vignette. Two different multi-run
functions exist for two different problems:

1. Perform Monte-Carlo simulations with single parameters set up for variation, and
2. simulations over multiple locations, parameter sets, or climate scenarios.

For the first case, the function `run_multi_LWFB90()` is available. The second
problem can be tackled using the function `run_multisite_LWFB90()`. 

Both functions are wrapper functions for `run_LWFB90()` and allow for parallel
processing of tasks, using a specified number of CPUs to speed up the execution
of a multi-run simulation. They return large lists containing the individual
single run simulation results, as they are returned by `run_LWFB90()`. These
result-lists can become very large if many simulations are performed, and the
selected output comprises daily data sets and especially the individual soil
layers' daily soil moisture states. Huge amounts of data produced can overload
the memory, I therefore start this vignette with a section on how to make best
use of the `output_fun`-argument of `run_LWFB90()` for output data management.

```{r, warning = FALSE, message = FALSE, eval = TRUE}
library(LWFBrook90R)
library(data.table)
```

## Data management (i): `output_fun`-argument

To minimize memory allocation, it is recommended to reduce the selected output
to a minimum and make use of the `output_fun`-argument of `run_LWFB90()`. With
this argument, it is possible to pass custom functions to `run_LWFB90()`, which
directly perform on the simulation output list object. With `rtrn_output =
FALSE`, the original simulation output (as selected by `output`) then can be
discarded, and only the results from the `output_fun`-argument are returned.
This can be very useful for model calibration or sensitivity analyses tasks
comprising ten thousands of simulations in a Monte-Carlo setting. With this
magnitude, memory allocation is critical, and only a relatively small output can
be returned for each individual simulation (e.g., a measure of agreement between
simulated and observed values). Similarly, it is possible to define functions
for custom output aggregation, to redirect the simulation output to a file or
database, as we will see later.

To demonstrate the usage of the `output_fun`-argument, we perform a Monte-Carlo
simulation using the function `run_multi_LWFB90()`, and define a function that
returns annual mean soil water storage and transpiration during the growing
season. In a first step, the function integrates depth-specific soil moisture to
soil water storage down to specified soil layer (`tolayer`), and in a second
step calculates mean soil water storage over the growing season, along with the
sum of transpiration. The growing season thereby is defined by the input
parameters `budburstdoy` and `leaffalldoy`.

```{r}
output_function <- function(x, tolayer) {
  # aggregate SWAT
  swat_tran <- x$SWATDAY.ASC[which(nl <= tolayer), 
                             list(swat = sum(swati)),
                             by  = list(yr, doy)]
  #add transpiration from EVAPDAY.ASC
  swat_tran$tran <- x$EVAPDAY.ASC$tran
  
  # get beginning and end of growing season from input parameters
  vpstart <- x$model_input$param_b90$budburstdoy
  vpend <- x$model_input$param_b90$leaffalldoy
  swat_tran <- merge(swat_tran,
                     data.frame(yr = unique(swat_tran$yr),
                                vpstart, vpend), by = "yr")
  # mean swat and tran sum
  swat_tran[doy >= vpstart & doy <= vpend, 
            list(swat_vp_mean = mean(swat), tran_vp_sum = sum(tran)), by = yr]
}
```


```{r, eval = TRUE, echo = FALSE}
data("slb1_meteo")
data("slb1_soil")
soil <- cbind(slb1_soil, hydpar_wessolek_tab(texture = slb1_soil$texture))
b90res <- LWFBrook90R:::b90res
```

To test our custom output function we run a single-run simulation

```{r}
data("slb1_meteo")
data("slb1_soil")
soil <- cbind(slb1_soil, hydpar_wessolek_tab(texture = slb1_soil$texture))
b90res <- run_LWFB90(options_b90 = set_optionsLWFB90(),
                     param_b90 = set_paramLWFB90(),
                     climate = slb1_meteo,
                     soil = soil,
                     output = set_outputLWFB90())
```

and apply the function to the return, to see that our custom output function works:

```{r}
output_function(b90res, tolayer = 15)
```

## Multi-run simulations with `run_multi_LWFB90()`

As mentioned, `run_multi_LWFB90()` is a wrapper for `run_LWFB90()`.
`run_multi_LWFB90()` takes a data.frame `paramvar` containing variable parameter
values in columns and their realizations in rows. For each row in `paramvar`,
the respective parameter values in `param_b90` are replaced by name, and
`run_LWFB90()` is called. Further arguments to `run_LWFB90()` have to be
specified and are passed on.

For the multi-run simulation, we set up two parameters for variation, the
maximum leaf area index (`maxlai`) and the maximum leaf conductance (`glmax`).
We define a data.frame with two columns, containing 50 random uniform
realizations of the two parameters:

```{r}
set.seed(2021)
N=50
paramvar <- data.frame(maxlai = runif(N, 4,7),
                       glmax = runif(N,0.003, 0.01))
```

Now we can run the simulation. We suppress the selected simulation result
objects and model input from being returned, and only return the values from our
`output_fun` defined above. We pass `tolayer = 15` so that soil water storage is
integrated down to the 15th soil layer, corresponding to 0-100 cm soil depth.
Note that the `param_b90` object (and thus parameters `budburstdoy` and
`leaffalldoy`) is available to our `output_fun`, although it is not included in
the return (`rtrn_input = FALSE`).

```{r}
mrun_res <- run_multi_LWFB90(paramvar = paramvar,
                             param_b90 = set_paramLWFB90(),
                             cores = 5, # arguments below are passed to run_LWFB90()
                             options_b90 = set_optionsLWFB90(), 
                             climate = slb1_meteo,
                             soil = soil,
                             output = set_outputLWFB90(), 
                             rtrn_input = FALSE, rtrn_output = FALSE,
                             output_fun = output_function,
                             tolayer = 15) # argument to output_fun
```

The result is a list of the individual single-run results, from which we can
easily extract the results of our output function and `rbindlist` them together
in a data.table:

```{r}
mrun_dt <- rbindlist(lapply(mrun_res, function(x) x$output_fun[[1]]), 
                     idcol = "singlerun")
```

```{r, eval = TRUE, echo = FALSE}
mrun_dt <- LWFBrook90R:::mrun_dt
```

Now we can display the results of the 50 simulations using boxplots:

```{r,results='hide', fig.height=5, fig.width=7, echo =FALSE, fig.cap = "Growing season soil water storage and transpiration of 50 simulations, with random variation of parameters 'maxlai' and 'glmax'"}

par(mfrow = c(1,2))
boxplot(swat_vp_mean~yr, data = mrun_dt, col = "blue")
boxplot(tran_vp_sum~yr, data = mrun_dt, col = "green")
```

We ran 50 simulations, all with the same climate, soil, and parameters except
for `maxlai` and `glmax`, that where varied randomly. In the next section we
will learn how to make use of multiple climate, soil, and parameter sets using
the function `run_multisite_LWFB90()`.

## Multi-Site simulations using `run_multisite_LWFB90()`

A basic multi-site simulation with the function `run_multisite_LWFB90()` runs
through lists of `param_b90`, `climate`, and `soil`-objects, and evaluates the
specified parameter sets for each of the soil/climate combinations. To
demonstrate its usage, we define two parameter sets, that we want to run on
three 'sites'. We include the two parameter sets in a list `parms_l`:

```{r}
parms_beech <- set_paramLWFB90(maxlai = 6)
parms_spruce <- set_paramLWFB90(maxlai = 4.5, winlaifrac = 0.8)
parms_l <- list(beech = parms_beech, spruce = parms_spruce)
```

We pretend that the three sites all have individual climates and soils, and set
up lists for soil and climate input:
```{r}
soils_l <- list(soil1 = soil, soil2 = soil, soil3 = soil)
climates_l <- list(clim1 = slb1_meteo, clim2 = slb1_meteo, clim3 = slb1_meteo)
```

Now we can run a small example: 
```{r}
msite_run1 <- run_multisite_LWFB90(
  options_b90 = set_optionsLWFB90(startdate = as.Date("2002-06-01"), 
                                  enddate = as.Date("2002-06-30")),
  param_b90 = parms_l,
  climate = climates_l,
  soil = soils_l,
  cores = 3)
str(msite_run1, max.level = 1)
```

The results are returned as a list of single run objects, with their names being
concatenated from the names of the input list entries holding the individual
`param_b90`, `climate`, and `soil` input objects.

## Data management (ii): A function as `climate`-argument

The function `run_multisite_LWFB90()` can easily be set up to run a few dozens
of sites with individual climate data, however, simulating thousands of sites
potentially is impossible, because such a large list of `climate` data.frames
might overload the memory of a usual desktop computer. Fortunately, it is
possible to pass a function instead of a data.frame as `climate`-argument to
`run_LWFB90()`. Such a function can be used to create the `climate`-data.frame
from a file or database-connection within `run_LWFB90()` or
`run_multisite_LWFB90()` on the fly. For `run_LWFB90()`, we can simply provide
arguments to the function via the `...`-placeholder. For
`run_multisite_LWFB90()`, we need to pass arguments to a `climate`-function
(possibly with individual values for individual site, e.g. a file name) via the
`climate_args`-argument.

To demonstrate this mechanism, we write three files with climatic data to a temporary location, from where we will read them back in later:

```{r}
tdir <- tempdir()
fnames <- paste0(tdir, "/clim", 1:3, ".csv")
write.csv(slb1_meteo[year(slb1_meteo$dates) %in% c(2002,2003),], file = fnames[1], row.names = FALSE)
write.csv(slb1_meteo[year(slb1_meteo$dates) %in% c(2002,2003),], file = fnames[2], row.names = FALSE)
write.csv(slb1_meteo[year(slb1_meteo$dates) %in% c(2002,2003),], file = fnames[3], row.names = FALSE)
```

For testing, we perform a single run with `runLWFB90()` and use the `fread`
function from the 'data.table'-package as `climate`-argument. The function reads
'csv'-files, and takes a `file` name as argument that we include in the call. It
points to the first of our three climate files:

```{r}
srun <- run_LWFB90(
  options_b90 = set_optionsLWFB90(),
  param_b90 = set_paramLWFB90(),
  soil = soil,
  output = set_outputLWFB90(),
  climate = fread,
  file = fnames[1],
  rtrn.input = FALSE)
```

The same construct basically works with the function `run_multisite_LWFB90()`.
The only difference to single-run simulation is that the arguments for the
function have to be specified in a named list of lists, one sub-list for each
site. We set it up as follows:
```{r}
clim_args <- list(climfromfile1 = list(file = fnames[1]),
                  climfromfile2 = list(file = fnames[2]),
                  climfromfile3 = list(file = fnames[3]))
```

Now we call `run_multisite_LWFB90()`, and set as `climate`-object the function
`fread`. Our list lists with individual arguments for `fread` is passed to the
function via `climate_args`:
```{r}
msite_run2 <- run_multisite_LWFB90(
  options_b90 = set_optionsLWFB90(),
  param_b90 = parms_l,
  soil = soils_l,
  climate = fread,
  climate_args = clim_args,
  cores = 3)

str(msite_run2, max.level = 1)
``` 

We simulated two parameter sets using three different climate/soil combinations.
The names of the climate used in the result names are now coming from the names
of our list `clim_args`, because we used a function as `climate`-argument. The
function `fread` is evaluated directly within `run_multisite_LWFB90()`, and is
not passed to `run_LWFB90()`, because otherwise it would have been evaluated for
each single-run simulation. In this way, `fread` is evaluated only three times
for in total six simulations which saves us some execution time, if we want to
simulate multiple parameter sets using the same climatic data.

## Multi-site simulation: Input from file, output to file

Now that we learned how to use a function as climate input, we can combine this
input facility with an `output_fun` that writes the simulation results to a
file. To do so, we extend our output function from the first example so that it
writes the results to a file in a specified directory. The file name is
constructed from the names of the current soil, climate, and parameter object,
which are passed automatically from `run_multisite_LWFB90()` to `run_LWFB90()`
as character variable `soil_nm`, `clim_nm`, and `param_nm`. In this way, the names of
currently processed input objects are accessible to `output_fun`-functions
within `run_LWFB90()`.

```{r}
output_function2 <- function(x, tolayer, basedir = getwd(),
                            soil_nm, clim_nm, param_nm ) {
  # file-name
  filenm = file.path(basedir, paste(soil_nm, clim_nm, param_nm, sep = "_"))
  
  # aggregate SWAT
  swat_tran <- x$SWATDAY.ASC[which(nl <= tolayer), 
                             list(swat = sum(swati)),
                             by  = list(yr, doy)]
  #add transpiration from EVAPDAY.ASC
  swat_tran$tran <- x$EVAPDAY.ASC$tran
  
  # get beginning and end of growing season from input parameters
  vpstart <- x$model_input$param_b90$budburstdoy
  vpend <- x$model_input$param_b90$leaffalldoy
  swat_tran <- merge(swat_tran,
                     data.frame(yr = unique(swat_tran$yr),
                                vpstart, vpend), by = "yr")
  # mean swat and tran sum
  swattran_vp <- swat_tran[doy >= vpstart & doy <= vpend, 
            list(swat_vp_mean = mean(swat), tran_vp_sum = sum(tran)), by = yr]
  
  write.csv(swattran_vp, file = paste0(filenm, ".csv"))
}
```

Now we can run the simulations, with climate data coming from files, and the results being written to our temporary directory `tdir`:  

```{r}
msite_run3 <- run_multisite_LWFB90(
  options_b90 = set_optionsLWFB90(),
  param_b90 = parms_l,
  soil = soils_l,
  climate = fread,
  climate_args = clim_args,
  output = set_outputLWFB90(),
  rtrn_input = FALSE, rtrn_output = FALSE,
  output_fun = output_function2,
  tolayer = 15,
  basedir = tdir,
  cores = 3)
```

After the simulation has finished, we can list the files and see that our attempt was successful:

```{r}
list.files(tdir, pattern = "csv")
```

```{r, echo = FALSE, results='hide'}
file.remove(list.files(tdir, pattern = "csv", full.names = T))
```



